A tool set for random number generation on GPUs in R

### 2021-11-27 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.Rnw
update paper1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3007 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

% $U=L*D^{1/2}*Z$, where 
% \begin{itemize}
% \itemsep0em 
% \item $L$ is the lower unit triangular matrix in Cholesky decomposition of $\Sigma = L*D*L^\top$,
% \item $D$ is a diagonal matrix.
% \item $Z=(Z_1, Z_2, \dots, Z_n) \sim \text{MVN}(0,I_n)$, where $I_n$ is a $n \times n$ identity matrix.
% \end{itemize}
The matrix decomposition method is straightforward to implement, however, when a large number of locations ($n$) is involved, the cost of Cholesky decomposition of the covariance matrix is $\mathcal{O}(n^3)$,  and the cost of matrix-vector multiplication $L*Z$ to generate the random field $U$ is $\mathcal{O}(n^2)$ \citep{LiuandLi2019}.  \pkg{clrng} package not only does the matrices decomposition (Cholesky decomposition and matrix-vector multiplication) in parallel, but also computes the covariance matrices in parallel on GPU. The following shows an example, in which we simulate 10 Gaussian random fields of mat\'ern covariance with 5 sets of parameters at one time. %and introduces several GPU functions used in the simulation process.

>>>>

% $U=L*D^{1/2}*Z$, where 
% \begin{itemize}
% \itemsep0em 
% \item $L$ is the lower unit triangular matrix in Cholesky decomposition of $\Sigma = L*D*L^\top$,
% \item $D$ is a diagonal matrix.
% \item $Z=(Z_1, Z_2, \dots, Z_n) \sim \text{MVN}(0,I_n)$, where $I_n$ is a $n \times n$ identity matrix.
% \end{itemize}
The matrix decomposition method is straightforward to implement. However, when a large number of locations $n$ is involved, the cost of Cholesky decomposition of the covariance matrix is $\mathcal{O}(n^3)$,  and the cost of matrix-vector multiplication $L*Z$ to generate the random field $U$ is $\mathcal{O}(n^2)$ \citep{LiuandLi2019}.  Our another \proglang{R} package \pkg{gpuBatchMatrix} is able to compute batches of Mat\'ern covariance matrices in parallel, and do Cholesky decomposition and matrix-matrix multiplication in batches in parallel on GPU. The following shows an example, in which we simulate on GPU 10 Gaussian random fields of mat\'ern covariance at one time with 5 sets of parameters, by taking advantage of the GPU-capabilities provided by \pkg{clrng} and \pkg{gpuBatchMatrix} together.
%\pkg{clrng} package not only does the matrices decomposition (Cholesky decomposition and matrix-vector multiplication) in parallel, but also computes the covariance matrices in parallel on GPU.

### 2021-11-27 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.tex
update paper1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3007 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is around 120 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streams <- createStreamsGpu(n = 512 * 128)
system.time(clrng::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.072   0.000   0.072
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(matrix(stats::rnorm(10000^2),10000,10000))
##    user  system elapsed 
##   6.658   1.125   7.779
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
week_GPU<-gpuR::vclMatrix(week,type="integer")
system.time(result_week<-clrng::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = c(256,64)))
##    user  system elapsed 
##   1.950   0.001   1.952
result_week$threshold
## [1] -54989.56
result_week$simNum
## [1] 10010624
result_week$counts
## [1] 1193
result_week$p.value
## [1] 0.0001192733
\end{verbatim}
\end{kframe}
\end{knitrout}
The ``week'' table has a much smaller p-value: around 0.0001249672, which should require a larger number of simulations to get a more accurate p-value. With more than ten million simulations, we get 1205 cases and a p-value around 0.000120472 by \fct{fisher.sim}. \fct{stats::fisher.test} takes 91.6 seconds, while \fct{fisher.sim} takes about 4.282 seconds, the elapsed time is decreased to about 4.67\% of the time taken on CPU.

>>>>

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is around 120 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streams <- createStreamsGpu(n = 512 * 128)
system.time(clrng::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.070   0.002   0.073
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(matrix(stats::rnorm(10000^2),10000,10000))
##    user  system elapsed 
##   6.658   1.125   7.779
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
week_GPU<-gpuR::vclMatrix(week,type="integer")
system.time(result_week<-clrng::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = c(256,64)))
##    user  system elapsed 
##   1.934   0.016   1.949
result_week$threshold
## [1] -54989.56
result_week$simNum
## [1] 10010624
result_week$counts
## [1] 1193
result_week$p.value
## [1] 0.0001192733
\end{verbatim}
\end{kframe}
\end{knitrout}
The ``week'' table has a much smaller p-value: around 0.0001249672, which should require a larger number of simulations to get a more accurate p-value. With more than ten million simulations, we get 1205 cases and a p-value around 0.000120472 by \fct{fisher.sim}. \fct{stats::fisher.test} takes 91.6 seconds, while \fct{fisher.sim} takes about 4.282 seconds, the elapsed time is decreased to about 4.67\% of the time taken on CPU.
